---
title: "Project 2 -- Signature creation and iLINCS API"
runtime: shiny
output: html_document
---
For the codes, check "app.R".
This assignment builds upon the R/shiny class and expands the API example.

1. For the assignment use the dataset TCGA_breast_cancer_ERstatus_allGenes.tsv. You may choose to "debug" your code with a smaller dataset TCGA_breast_cancer_ERpositive_vs_ERnegative_PAM50.tsv or TCGA_breast_cancer_LumA_vs_Basal_PAM50.tsv. 
```{r setup,include=FALSE}
library(shiny)
library(DT)
library(httr)
library(ECharts2Shiny)
options(shiny.maxRequestSize=70*1024^2)
```


2. Your assignment is to develop code using R Shiny to present user interface that allows a user to upload a tsv file, presents an option to split the samples into two groups based on the available metadata and create a signature. The signature is then submitted to the iLincs API which retrieves concordant signatures.
```{r,include=FALSE}
ui <- fluidPage(
  
  # Application title
  titlePanel("File upload"),

  # Sidebar with a slider input for number of bins
  sidebarLayout(
    sidebarPanel(
      fileInput('file1', 'Choose TSV File', accept=c('text/tsv','.tsv')),   
      selectInput("variable", "Grouping Variable:", choices=c()),
      selectizeInput('group1', "Group1", choices = NULL, multiple = TRUE),   #you could select multiple values when using selectize
      selectizeInput('group2', "Group2", choices = NULL, multiple = TRUE),
      selectInput("difffunction", "Differential function:", choices=c("t-test")),
      checkboxInput("L1000",label = "Limit Genes to L1000",value=FALSE),
      selectInput("topgenes","Top Genes:",choices=c("All","Top 100")),
      actionButton("compute", "Compute and Submit Signature"),
      actionButton('getHmap','Get Heatmap')
    ),

    # Show a plot of the generated distribution
    mainPanel(
      dataTableOutput("signature_data"),
      hr(),
      dataTableOutput("correlated_data"),
      hr(),
      dataTableOutput("sample_data"),
      hr(),
      plotOutput("themap")
    )
  )
)
server <- function(input, output, session) {
  
  values <- reactiveValues(data=NULL)
  
  observe({    #create the reactive context
    # handle file upload 
    
    # input$file1 will be NULL initially. After the user selects
    # and uploads a file, it will be a data frame with 'name',
    # 'size', 'type', and 'datapath' columns. The 'datapath'
    # column will contain the local filenames where the data can
    # be found.
    
    inFile <- input$file1
    
    if (is.null(inFile))
      return(NULL)
    
    isolate({ 
      file <- (inFile$datapath)

      values$header <- scan(file, nlines = 1, sep="\t", what = character())
      values$data <- read.table(file, skip = 2, header = FALSE, sep = "\t", quote = "", check.names=FALSE)
      names(values$data) <- values$header 
      values$header2 <- data.frame(scan(file, skip = 1, nlines = 1, sep="\t", what = character()))
    })
  })
  
  # show sample metadata
  output$sample_data <- renderDataTable({
    head(values$data[,1:10],n=50)
  }, caption = "First 50 genes, first 10 samples")
  
  observe({
     # fill in variable selection from the input file
     updateSelectInput(session, "variable", choices=as.character(values$header2[1,]))
  })
  
  # handle UI group values update based on selected variable
  observe({
    if (input$variable !="") {
      
      updateSelectizeInput(session, 'group1', choices=unique(values$header2[-1,values$header2[1,]==input$variable]), server = TRUE)
      updateSelectizeInput(session, 'group2', choices=unique(values$header2[-1,values$header2[1,]==input$variable]), server = TRUE)
    }
  })
  
  # Create signature, upload to API, display results
  observeEvent(input$compute, {
    withProgress(message = 'Creating signature', value = 0, {
      
      #
      # Filter into two groups
      #
      group1 <- names(values$data)[values$header2==input$group1]
      group2 <- names(values$data)[values$header2==input$group2]

      #
      # ensure the values are numeric
      #
      values$values <- values$data[complete.cases(values$data), -1]
      values$values[] <- lapply(values$values, function(x) { as.numeric(as.character(x)) })
      
      #
      # select differential function
      #
      incProgress(1/3, detail = paste0("Running ",input$difffunction))
      if (input$difffunction=="t-test") {
        diff_result <- as.data.frame(apply(values$values, 1, 
                                           function(x) { t.test(unlist(x[group1], use.names = FALSE),
                                                              unlist(x[group2], use.names = FALSE))$p.value }))
        
        # FIXME: fill in
        # Add t statistic as the measure of differential expression (Value_LogDiffExp)
        diff_result$Value_LogDiffExp <- apply(values$values, 1, 
                                              function(x) { t.test(unlist(x[group1], use.names = FALSE),
                                                                 unlist(x[group2], use.names = FALSE))$statistic })
      }
      
      #
      # format signature output
      #
      output_id_column_name <- paste0(values$header[1],"_GeneSymbol")
      diff_result <- data.frame(values$data[values$header[1]], diff_result[,1:2])
      colnames(diff_result) <- c(output_id_column_name, "Significance_pvalue", "Value_LogDiffExp")
      
      diff_result <- diff_result[, c(1, 3, 2)]
                                              
      # FIXME: fill in
      # choose only L1000 genes
      if (input$L1000 == TRUE) {
        genes<-read.csv2("L1000.txt", sep='\t')
        L1000 <- genes[genes$pr_is_lm=='1',]$pr_gene_symbol
        diff_result <- diff_result[diff_result[,1] %in% L1000,]
      }
      
      # FIXME: fill in
      # choose top 100 most differentially expressed genes
      # diff_result <- ...
      if (input$topgenes == "Top 100") {
        diff_result <- head(diff_result[order(-abs(diff_result$Value_LogDiffExp)),],n=100)  #The most differentiated values are at the top
      }
      
      #
      # show signature in a table
      #
      output$signature_data <- DT::renderDataTable({
        diff_result
      }, caption = "Signature to submit to iLincs")
      
      incProgress(1/3, detail = paste("Submitting the signature to iLincs"))
      
      #
      # create temporary csv file to submit into API
      #
      ftemp <- tempfile(pattern = "file", fileext=".csv", tmpdir = tempdir())
      write.csv(diff_result, ftemp, row.names = F, quote=F)
      cat(ftemp)
      
      r <- POST("http://www.ilincs.org/api/SignatureMeta/uploadAndAnalyze?lib=LIB_5", body = list(file = upload_file(ftemp)))
      
      l <- lapply(content(r)$status$concordanceTable, function(x) unlist(x))
      ilincs_result <- data.frame(t(sapply(l,c)))
      
      #
      # show correlation results
      #
      output$correlated_data <- DT::renderDataTable({
        datatable( ilincs_result, rownames = TRUE, caption = "Correlated signatures")
      })
    })
  })
  
  #  create heat map of a file
  observeEvent(input$getHmap, {
    withProgress(message = 'Creating heatmap', value = 0, {
      #load the data
      file <- "TCGA_breast_cancer_ERpositive_vs_ERnegative_PAM50.tsv" 
      #prepare data
      header <- scan(file, nlines = 1, sep="\t", what = character())
      data <- read.table(file, skip = 2, header = FALSE, sep = "\t", quote = "", check.names=FALSE)
      names(data) <- header
      header2 <- scan(file, skip = 1, nlines = 1, sep="\t", what = character())
      row.names(data) <- data$id
      data <- data [,2:1037]
      data <- as.matrix(data)
      output$themap <- renderPlot({
        heatmap(data)
     })
    })
  })
}
shinyApp(ui = ui, server = server)
```

3. Your assignment is to develop code using R Shiny to present user interface that allows a user to upload a tsv file, presents an option to split the samples into two groups based on the available metadata and create a signature. The signature is then submitted to the iLincs API which retrieves concordant signatures.

4. Template is provided in the Project 2 github project (https://github.uc.edu/uc-datascience/Project2.git).

5. The template is missing the **calculation of the differential expression** -- please use **t-test** to calculate **t-statistic / p-value**.

6. Expand the template to allow users to **filter the input file to L1000 genes** only (See the include L1000.txt file).
7. Further expand the template to allow users to **submit only top 100 differentially expressed genes**.
8. **Compare results with iLincs**

**MY ANSWER:** The results are the same as what iLincs shows. For example, when I search raloxifene in both correlated signatures table of my result and LINCS chemical perturbagen signatures using the newly generated csv, I can find that the p-values and similarity of both exactly match. 

9. Extra credit for a **heatmap or other visualizations**.
(included above)
The assignment is due on -- March 23, 2020 midnight. 

The submission should be zip compressed file named "project2-[*your UC username*].zip" (e.g. "project2-lastnfi.zip") which includes any supporting R files. The zip file should be uploaded canopy. The assignment entry in Canopy will be created shortly.


